---
title: Accidental Drug Related Deaths
author: Kyle Harris
date: '2018-05-27'
categories: ["R"]
tags: ["opioids"]
---

```{r setup, include=FALSE}
# display 2 decimals
options(digits=3)

# data
odDat <- read.csv('/Users/kow/Box Sync/HDS/B202/Final/Accidental_Drug_Related_Deaths__2012-June_2017.csv')

# libraries
library(zoo)
library(stringr)
library(tidyverse)
library(lubridate)
library(TeachingDemos)
library(ggthemes)
library(kableExtra)
library(broom)
library(knitr)
```

# Follow Me On

<a href="https://twitter.com/KoderKow"> <i class='fab fa-twitter fa-3x'></i></a> <a href="https://www.instagram.com/koderkow/"> <i class='fab fa-instagram fa-3x'></i></a> <a href="https://github.com/KoderKow/"><i class='fab fa-github fa-3x'></i></a> <a href="https://koderkow.rbind.io"><i class='fas fa-globe fa-3x'></i></a>

# Intro

The opioid epidemic is a nationwide issue for the United States. Among the states Connecticut takes second place in terms of deaths from opioids in the years 2014-2015. As the years go on death rates are doubling from the abuse of opioids. Heroin overdoses continue to rise and now fentanyl is joining its ranks. Fentanyl has been getting mixed with heroin to increase its potency and in return it increases the chance of overdosing (Rondinone, 2017). This report will be looking over the top 5 opioids that are related to overdosing. The data used is from [data.ct.gov](https://data.ct.gov/Health-and-Human-Services/Accidental-Drug-Related-Deaths-2012-2017/rybz-nyjw). This project was done during a class at IUPUI in November 2017.

---

## Data Exploration

We loaded the data into a dataframe called 'odDat'. Our first step is to make more columns according to the date.

```{r}
odDat$yearMon <- as.yearmon(odDat$Date,'%m/%d/%Y')
odDat$yearMonNum <- as.numeric(odDat$yearMon)
odDat$year <- year(mdy(odDat$Date))
odDat$month <- month(mdy(odDat$Date))
odDat$day <- day(mdy(odDat$Date))
```

We will be looking into the location type for the drug over doses. Lets see a count of them.

```{r message=FALSE, warning=FALSE, results='hide'}
odDat %>%
  group_by(Location) %>%
  summarise(n = n()) %>%
  arrange(desc(n))
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
odDat %>%
  group_by(Location) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

Most of our data is within 3 locations. So we will be aggragating and grouping our data based off these 3 locations. We will be working with the dataset renamed as 'od'.

```{r}
od <- odDat %>%
  filter(Location == "Residence" |
           Location == "Hospital" |
           Location == "Other")
```

Lets look at an example of the data we are wanting to find out about. For this example we will be looking at heroin overdoses in locations marked as "Other."

```{r eval=FALSE}
od %>%
  filter(Location == "Other" & Heroin == "Y") %>%
  summarise(n = n())
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
od %>%
  filter(Location == "Other" & Heroin == "Y") %>%
  summarise(n = n()) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

There are 330 counts of heroin overdose in the location known as other. I want to clean the overdose data to perform better insights. Any case of 'Y' or 'y' will be marked as a 1, if it is not a Y then we will label it as a 0.

```{r}
od[,16:27] <- ifelse(od[,16:27] == "Y" | od[,16:27] == 'y',1, 0)
od[,29] <- ifelse(od[,29] == "Y" | od[,29] == 'y', 1, 0)
```

For this project we are interested in the top 5 drugs by count of overdoses. We will use dplyr to select the columns, and then summarize them to find the top 5 most common overdoses.

```{r}
drugs <- od %>%
  select_(.dots = c("Location", "Heroin", "Cocaine", "Fentanyl",
                    "Oxycodone", "Oxymorphone", "EtOH",
                    "Hydrocodone", "Benzodiazepine", "Methadone",
                    "Amphet", "Tramad", "Morphine..not.heroin.",
                    "Any.Opioid", "yearMon"))

drugCount <- drugs %>%
  summarise(Heroin = sum(Heroin),
            Cocaine = sum(Cocaine),
            Fenentanyl = sum(Fentanyl),
            Oxycodone = sum(Oxycodone),
            Oxymorphone = sum(Oxymorphone),
            Etoh = sum(EtOH),
            Hydrocodone = sum(Hydrocodone),
            Benzodiazepine = sum(Benzodiazepine),
            Methadone = sum(Methadone),
            Amphet = sum(Amphet),
            Tramad = sum(Tramad))
```

Display the top 10 opioids.

```{r eval=F}
drugCount %>%
  gather() %>%
  top_n(10) %>%
  arrange(desc(value))
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
drugCount %>%
  gather() %>%
  top_n(10) %>%
  arrange(desc(value)) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

The top 5 in order of highest to lowest is heroin, fentanyl, cocaine, benzodiazepine, and etOh. It is no surprise to see heroin and fentanyl are on top. Looking at the information for these opioids we are able to look into where the most common area for overdosing is to happen.

```{r results='hide'}

drugs5 <- od %>%
  select_(.dots = c("Location", "Heroin", "Cocaine",
                    "Fentanyl", "EtOH", "Benzodiazepine",
                    "yearMon"))

drugs5 %>% summarize(Residence = sum(Location == 'Residence'),
                     Hospital = sum(Location == 'Hospital'),
                     Other = sum(Location == 'Other'))
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
drugs5 %>%
  summarize(Residence = sum(Location == 'Residence'),
                     Hospital = sum(Location == 'Hospital'),
                     Other = sum(Location == 'Other')) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

Residence has the most occurances, followed by hospital then other. Now lets look at how location is distributed among the top 5 drugs. We will do this with two different bar plots. The first will be a normal bar plot, with each bar representing counts of location per drug.

```{r}
drugs5Loc <- drugs5 %>%
  group_by(Location) %>%
  summarize(nHeroin = sum(Heroin),
            nFentanyl = sum(Fentanyl),
            nCocaine = sum(Cocaine),
            nBenzodiazepine = sum(Benzodiazepine),
            nEtOh = sum(EtOH)) %>%
  gather(opioid, count, 2:6)

drugs5Loc$opioid <- str_replace(drugs5Loc$opioid, 'n', '')

ggplot(data = drugs5Loc,
       aes(x = reorder(opioid, count), y = count, fill = Location)) +
  geom_bar(stat = 'identity', position='dodge',
           color = 'black') +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = 'opioid', y = 'Count')
```

The other bar plot I will use is a proportion based bar plot so it is easier to see how location makes up each opiod.

```{r}
ggplot(data = drugs5Loc, aes(x = opioid, fill = Location, y = count)) +
  geom_bar(position = 'fill', stat = 'identity') +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ylab('Proportion') +
  xlab('opioid')
```

Looking at these graphs we can confirm that residence makes up most cases for each drug. Generally hospital and other are lower counts, with hospital being the next common and then other. Now that we have an understanding of the top opioids and locations that they occur, it is time to look into overdose counts over the months and years.

```{r message=FALSE}
yearMonDrug <- od %>%
  group_by(yearMon, Location) %>%
  summarize(Heroin = sum(Heroin),
            Cocaine = sum(Cocaine),
            Fentanyl = sum(Fentanyl),
            Etoh = sum(EtOH),
            Benzo = sum(Benzodiazepine)) %>%
  gather(opioid, count, 3:7)

yearMonDrug %>%
  ggplot(aes(x = yearMon, y = count, fill = opioid)) +
  geom_bar(stat = "identity") +
  facet_grid(opioid ~ Location) +
  labs(y = 'Count', x = 'Year') +
  scale_colour_gdocs() + theme_gdocs()
```

We can see that fentanyl has had a huge spike in overdoses in 2016. Before late 2015 fentanyl is one of the lowest overdosed drugs. This truly shines the light on how fentanyl is a major epidemic. There is a constant increase for all drugs in all locations with a very few observations of strong decreases.

```{r}
yearMonDrug %>%
  ggplot(aes(x = yearMon, y = count, fill = opioid)) +
  geom_bar(stat = "identity") + facet_wrap( ~ Location) +
  labs(x = "Year", y = "Count") +
  theme(axis.text.x  = element_text(angle=45)) + 
          scale_x_continuous(breaks=c(2012, 2013, 2014, 2015, 2016
                                      ,2017))
```

We can clearly see the constant increase of overdoses. Next we will facet the plot to look at each opioid individually.

```{r}
# plot to show count of top 5 drugs over 2012-2017
yearMonDrug %>%
  ggplot(aes(x = yearMon, y = count, fill = opioid)) +
  geom_bar(stat = "identity") + facet_wrap( ~ opioid) +
  labs(x = "Year", y = "Count") +
  theme(axis.text.x  = element_text(angle=45)) + 
          scale_x_continuous(breaks=c(2012, 2013, 2014, 2015, 2016
                                      ,2017))
```

Looking at all of the data we can see that the primary location for overdosing is residence. In all locations we can see that the count is increasing. Heroin and fentanyl are the most common as well. It is interesting to see the huge spike in fentanyl count over the past few years. The large increase is so much it now has a higher count than heroin.

---

## Simulation

Moving on past visual representation and aggragation of our data we will now look into the last 6 months of our data set, more specifically the first 6 months of 2017. I will be using these 6 months to simulate the next few months that are not present in our data. Janurary overdose count will be used for July, Feburary for August, ect. I Will be simulated the first 3 months for each drug. Then I will break down the simulation by location for the first three months, then I am interested in doing heroin, the most popular opioid, by location for the last 6 months.

This process may not make sense logically. This was a requirement by the project to implement simulation. It is moreso theory in how to apply simulation in this kind of setting and how to interpret it.

```{r}
set.seed(123)

od2 <- od %>%
  select(Heroin, Cocaine, Fentanyl, EtOH,
         Benzodiazepine, yearMon, year, month, Location)

# values for overdoses in Jan 2017
od20171 <- od2 %>%
  group_by(Location) %>%
  filter(year == 2017 & month == 1) %>%
  summarize(nHeroin = sum(Heroin),
            nCocaine = sum(Cocaine),
            nFentanyl = sum(Fentanyl),
            nEtoh = sum(EtOH),
            nBenzo = sum(Benzodiazepine))
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
od20171 %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

This shows us the counts of opioids in Janurary 2017.

```{r}
# sim for july
set.seed(char2seed('Kyle'))
# sims for each drug by total
july2017 <- data.frame(Heroin=rpois(1000,sum(od20171$nHeroin)),
                       Cocaine=rpois(1000,sum(od20171$nCocaine)),
                       Fent=rpois(1000,sum(od20171$nFentanyl)),
                       Etoh=rpois(1000,sum(od20171$nEtoh)),
                       Benzo=rpois(1000,
                                sum(od20171$nBenzo)))
july2017 %>%
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_histogram(color='black',binwidth = 2,aes(fill=key)) +
  theme(legend.position="none") +
  labs(x="opioid", y="Overdose Count", title = "July 2017 Simulation")
```

This is five separate simulations for each drug to show, in a simulated manner, the count of overdoses for July 2017 based off the counts of Janurary 2017. Taking information from these simulations we can see that Benzo will have around 10 overdoses on the low end and we can expect a high of around 30 (10-30). Following this template, cocaine will have 20-42, Etoh 5-17, fentanyl will have 35 - 70, and heroin 31 - 65. Looking at summaries of our data before it is to be expected heroin would be higher. I will do a few more simulations for August and September to see if fentanyl is on top and to see if our simulated results show an increase/decrease to other.

To avoid summarizing each set, I will add the results of observations into a data frame and print it out.

```{r echo=F}
simRange <- data.frame(opioid = c("Benzo", "Cocaine", "Etoh", "Fentanyl", "Heroin"), Range = c('10-30', '20-42', '5-17', '35-70', '41-65'), Month = c("Jul", "Jul", "Jul", "Jul", "Jul"))
```

```{r}
od20172 <- od2 %>%
  group_by(Location) %>%
  filter(year == 2017 & month == 2) %>%
  summarize(nHeroin = sum(Heroin),
            nCocaine = sum(Cocaine),
            nFentanyl = sum(Fentanyl),
            nEtoh = sum(EtOH),
            nBenzodiazepine = sum(Benzodiazepine))

set.seed(char2seed('Kyle'))
# sims for each drug by total
august2017 <- data.frame(Heroin=rpois(1000,sum(od20172$nHeroin)),
                       Cocaine=rpois(1000,sum(od20172$nCocaine)),
                       Fent=rpois(1000,sum(od20172$nFentanyl)),
                       Etoh=rpois(1000,sum(od20172$nEtoh)),
                       Benzo=rpois(1000,
                                sum(od20172$nBenzodiazepine)))
august2017 %>%
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_histogram(color='black',binwidth = 2,aes(fill=key)) +
  theme(legend.position="none") +
  labs(x="opioid", y="Overdose Count", title = "August 2017 Simulation")
```

```{r}
od20173 <- od2 %>%
  group_by(Location) %>%
  filter(year == 2017 & month == 3) %>%
  summarize(nHeroin = sum(Heroin),
            nCocaine = sum(Cocaine),
            nFentanyl = sum(Fentanyl),
            nEtoh = sum(EtOH),
            nBenzodiazepine = sum(Benzodiazepine))

set.seed(char2seed('Kyle'))
# sims for each drug by total
sept2017 <- data.frame(Heroin=rpois(1000,sum(od20173$nHeroin)),
                       Cocaine=rpois(1000,sum(od20173$nCocaine)),
                       Fent=rpois(1000,sum(od20173$nFentanyl)),
                       Etoh=rpois(1000,sum(od20173$nEtoh)),
                       Benzo=rpois(1000,
                                sum(od20173$nBenzodiazepine)))
sept2017 %>%
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_histogram(color='black',binwidth = 2,aes(fill=key)) +
  theme(legend.position="none") +
  labs(x="opioid", y="Overdose Count", title = "September 2017 Simulation")
```

```{r echo = F}
simRange <- data.frame(opioid = c("Benzo", "Cocaine", "Etoh", "Fentanyl", "Heroin", "Benzo", "Cocaine", "Etoh", "Fentanyl", "Heroin", "Benzo", "Cocaine", "Etoh", "Fentanyl", "Heroin"),
                  Month = c("Jul", "Jul", "Jul", "Jul", "Jul", "Aug", "Aug", "Aug", "Aug", "Aug", "Sept", "Sept", "Sept", "Sept", "Sept"),
                  Range = c('10-30', '20-42', '5-17', '35-70', '41-65', '12-36', '15-42', '0-17', '30-66', '22-50', "15-40", "18-44", "0-20", "37-76", "27-56"))
```

```{r results='hide'}
simRange %>%
  filter(opioid %in% c("Fentanyl", "Heroin", "Benzo",
                                   "Cocaine", "Etoh")) %>%
  arrange(opioid)
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
simRange %>%
  filter(opioid %in% c("Fentanyl", "Heroin", "Benzo",
                                   "Cocaine", "Etoh")) %>%
  arrange(opioid) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

To conclude our simulations for the next three months we can see that benzo overdoses are on the rise. Cocaine is roughly constant. Etoh is also constant and is the lower overdose count. Fentanyl seems to be the most common overdose in 2017, it has a large range and the highest values. Heroin is now the second most common in these 2017 simulations, and it appears there is a big drop off from 41 to the 20s of overdose counts.

To finalize the simulations I will simulate July 2017 for heroin based off of location.

```{r results='hide'}
locationSim <- od2 %>%
  group_by(Location) %>%
  filter(year == 2017 & month == 1) %>%
  summarize(nHeroin = sum(Heroin))
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
locationSim  %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

```{r}
julyLocSim <- data.frame(Hospital = rpois(1000,
                                        locationSim$nHeroin[1]),
                         Other = rpois(1000,
                                         locationSim$nHeroin[2]),
                         Residence = rpois(1000,
                                        locationSim$nHeroin[3]))

julyLocSim %>%
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_histogram(color='black',binwidth = 2,aes(fill=key)) +
  theme(legend.position="none") +
  labs(x="Location", y="Overdose Count", title = "July 2017 Location Simulation")
```

The distribution of these poisson simulations look normal. This is because they have a smaller lamba due to the total count for the month being split into three separate locations. For heroin count in the hospital in July 20717 is no less than 8 and expect a high of 26. For ther it is 4 to 23, and 10 to 37 for residence. From this breakdown we are able to confirm what was observed earlier with our data exploration. According to our simulations residence is the most common location, followed by hospital, then other. I wanted to show a simulation breakdown according to location for a particular drug. This can be applied to any drug of interest.

---

## Predictive Modeling

Making a model for this dataset will require aggregating the data and getting the total counts by month and year. This is because we will be using a poisson model and it will require total counts. If we try to use the base data set we will not get the results we are expecting because opioids are composed of 0's and 1's. The training set, all data before 2017, will be used to train our model. For this example we will be looking specifically at the opioid fentanyl. Then I will test the model against a test set, all data after 2017. At the end the error will be calculated to show us the accuracy of the model.

```{r model}
trainod <- od2 %>%
  filter(year < 2017) %>%
  group_by(Location, year, month) %>%
  summarize(nHeroin = sum(Heroin),
            nCocaine = sum(Cocaine),
            nFentanyl = sum(Fentanyl),
            nEtoh = sum(EtOH),
            nBenzodiazepine = sum(Benzodiazepine))

model1 <- glm(nFentanyl ~ Location + year + month, data = trainod,
            family = 'poisson')
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
tidy(model1)  %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

The model, model1, will be used now predict the first month of 2017. Showing the actual data and the predicted data will show if the model is predicting close to the real data.

```{r}
new.predict1 <- predict(model1,newdata = data.frame(year=2016, Location = c('Hospital', 'Other', 'Residence'), month=12), type = "response")
names(new.predict1) <- c('Hospital','Other','Residence')
```

```{r results='hide'}
# predicted values for Jan 2017 in all locations for fentanyl
new.predict1
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
data.frame(as.list(new.predict1))  %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

Actual values for Jan 2017 in all locations for fentanyl.

```{r results = 'hide'}
od2 %>%
  filter(year == 2017 & month == 1) %>%
  summarize(nFentanyl = sum(Fentanyl))
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
od2 %>%
  filter(year == 2017 & month == 1) %>%
  summarize(nFentanyl = sum(Fentanyl)) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c('condensed', 'bordered'), full_width = F, position = 'center')
```

As shown after adding up the three individual predictions our predicted values are very close to the actual data. Exciting! To test the accuracy of the model I will now run the model against the test data, which is data in 2017. This way we will be able to gauge the accuracy of our model.

```{r}
testod <- od2 %>%
  filter(year < 2017) %>%
  group_by(Location, year, month) %>%
  summarize(nHeroin = sum(Heroin),
            nCocaine = sum(Cocaine),
            nFentanyl = sum(Fentanyl),
            nEtoh = sum(EtOH),
            nBenzodiazepine = sum(Benzodiazepine))

predTest <- predict(model1, newdata = testod, type = "response") 

errTest <- mean((testod$nFentanyl - predTest) ^ 2)

sqrt(errTest)
```

The model has a 2.5 error, meaning our predications are potentially off by 2.5. I believe this is accurate for our model. This could be very useful to predict upcoming months for the overdoses. The model is for heroin, but it can easily be switched with any of the other drugs in the data set.

To summarize our findings there is definite proof that overdose counts are on the rise. Residences are common locations for these accidents. Heroin is the most common in the data set, but it is slowly getting over taken by fentanyl. We are able to simulate ranges for expected overdoses in the data. This could be used for professionals to get an idea what they are looking for in the future. Finally we built a fairly accurate model. This model is able to take in year, month, and location and return a value that represents the count of overdoses. I believe this model could be useful to see if counts increase in the future. Furthermore after more data is acquired it would be great to see predicted counts go down because this would show overdose situations are under control.

---

## Sources + Related Links

- [Nicholas Rondinone, CDC: Connecticut Second In Percent Increase Of Synthetic Opioid Deaths Rate](http://www.courant.com/breaking-news/hc-cdc-fentanyl-overdoses-20170105-story.html)
- [Jeff Mill, Sen. Murphy in Haddam on opioid crisis: ‘This is getting worse, not better’](https://www.middletownpress.com/news/article/Sen-Murphy-in-Haddam-on-opioid-crisis-This-12548669.php)
- [Robert Storace, Attorneys for CT Cities Pledge Fight Against Move to Opioid MDL](https://www.law.com/ctlawtribune/sites/ctlawtribune/2018/01/24/attorneys-for-connecticut-cities-pledge-to-fight-any-move-to-opioid-mdl/)
- [Russel Blair, Eight New Laws That Take Effect in Connecticut on Jan. 1](http://www.courant.com/politics/hc-pol-new-connecticut-laws-january-20171227-story.html)

<!-- ```{r echo=FALSE} -->
<!-- blogdown::shortcode('googleAdsense') -->
<!-- ``` -->
