---
title: Tesseract OCR Version 4.1.0 Parameters
author: Kyle Harris
date: '2019-12-04'
slug: tesseract-ocr-version-4-1-0-parameters
categories:
  - R
tags:
  - Tesseract
  - NLP
  - Image Processing
  - Parameter
description: ~
featured_image: ~
---

```{r setup, include=FALSE}
library(vroom)
library(jsonlite)
library(dplyr)
library(DT)
library(kowr)

## Tesseract Data
url <- "https://raw.githubusercontent.com/KoderKow/stuff/master/data/tp_json.json"
tesseract_parameters <- fromJSON(url)
```

# Overview

During a project my team was having a hard time tracking down an up to date parameters list for [Tesseract OCR](https://github.com/tesseract-ocr/tesseract). The documentation mentions to use <hc>tesseract --print-parameters</hc> (two hyphens in front of print) to get the available *600+* options. Personally, I use R at my job so sifting through the terminal and trying to go through a poorly formatted (this may be due to my unfamiliarity to the terminal) list in the terminal can be a headache. I have taken the current parameters list for **Tesseract OCR 4.1.0** and made it available on this page and in JSON format in hopes it will help people explore the available options.

<!-- Below the table there will will be this table is JSON for those wanting the raw data. -->

# Cleaning Process

I used RStudio's find/replace functionality to clean up the text I copied from the terminal using the <hc>tesseract --print-parameters</hc> command. The *Description* column has apostrophe's **removed** due to the major conflicts it was causing me during data cleaning. There are some single and double quotes in the *Value* column, these were intially removed and then re-added. I have triple-checked to ensure they are correct since *Value* is important when looking at the default value of these parameters.

Thank you Tesseract OCR dev team for this fantastic piece of software and making it open source! :)

## JSON Source

If you are interested in the data used in the table, [here](https://raw.githubusercontent.com/KoderKow/stuff/master/data/tp_json.json) is a link to the raw data on [my GitHub](https://github.com/KoderKow/stuff/blob/master/data/tp_json.json).

# Parameters

```{r table, echo=FALSE}
tesseract_parameters %>% 
  snake_to() %>% 
  datatable(
    options = list(
      dom = 'ft',
      ordering = FALSE,
      pageLength = 650
      ),
    rownames = FALSE
    )
```
